---
title: "Cluster"
author: "Hannah Carney"
date: "2025-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project is analysing trends in population change in different counties across the US, including the overall increases/decreases in population and the composition of those population changes, ie. is there an increase in married or single households? Analysis of the composition of such population changes reveals socioeconomic factors and whether these influence the changes in population trends and composition. Finally, this analysis clusters counties based on their popualtion trends to identify similarities in population evolution over the past ten to fifteen years.

A problem that I found in this report is the student's decision to fully exclude LA county from the analysis. This is both a non-technical and a technical error. 

Non-technical: Although excluding LA may be reasonable given its outlying properties in terms of population size and social factors, the student gives little evidence and explanation of this in the report. It seems like they just say 'LA is on a cluster on it;s own so it must be distorting the other data' without providing any logical proof or explanation as to why. This logic also ignores the fact that outliers can provide valuable information. The fact that LA is an outlier is not only a source of distortion, but it also reveals interesting information about why this county is different from the other ones. The fact that it is in a cluster on its own may actually provide insight for the analysis as it allows us to identify what features distinguish it from other counties. This revised logic leads to the implementation of my technical improvement.

Technical: The simple improvement I made was commenting out the removal of LA county from the scaled df before k means clustering. I then made further technical and graphical edits to the code to clearly demonstrate the outlying properties of LA. I performed k means clustering as before, with 4 centers, however I used the graph to identify the unique properties of LA by colouring it separately from the other clusters (in green), in other words the graph is color coded by cluster, and then a separate color is applied to any point belonging to LA. I did the same thing for the t-sne analysis.

```{r}
library(glue)
library(Rtsne)
library(RColorBrewer)

df <- read.csv("../python/curve_feats_counties.csv")
df <- df[, -which(grepl('2023', names(df)))]
ts <- read.csv("../python/with_geo_household_cnt.csv")

scaled_df <- apply(df[, -1], 2, scale, center=TRUE, scale=TRUE)
scaled_df <- as.data.frame(scaled_df)
scaled_df[['NAME']] <- df$X
head(scaled_df)

scaled_df[['no_curve_married']] <- (is.na(scaled_df[['married.slope_2022']]) - 0.5) * 2
scaled_df[['no_curve_unmarried']] <- (is.na(scaled_df[['unmarried.slope_2022']]) - 0.5) * 2
for(i in seq_len(ncol(scaled_df))){
  is_na_vals <- is.na(scaled_df[, i])
  scaled_df[is_na_vals, i] <- 0 # replaced with mean, just so kmeans will run!
}

no_m_yes_u <- (scaled_df[['no_curve_married']] == 1) & (scaled_df[['no_curve_unmarried']] == -1)
yes_m_no_u <- (scaled_df[['no_curve_married']] == -1) & (scaled_df[['no_curve_unmarried']] == 1)
no_m_no_u <- (scaled_df[['no_curve_married']] == 1) & (scaled_df[['no_curve_unmarried']] == 1)
scenarios <- list(no_m_yes_u, yes_m_no_u, no_m_no_u)
png('look_at_bad_fits.png', 1200, 500)
par(mfrow=c(1, 3))
for(i in seq_along(scenarios)){
    s <- scenarios[[i]]
    rand_county <- sample(scaled_df[s, "NAME"], 1)
    sdf <- ts[ts$NAME == rand_county, c('year', 'B11002_003E', 'B11002_012E')]
    y_range <- c(min(sdf[, -1]), max(sdf[, -1]))
    plot(sdf$year, sdf[, 2], col="blue", pch=16, ylim=y_range)
    points(sdf$year, sdf[, 3], col="red", pch=16)
    legend("topleft", legend=c('married', 'unmarried'),
           fill=c('blue', 'red'))
}
dev.off()

#scaled_df <- scaled_df[scaled_df$NAME != 'Los Angeles County, California', ]
ks <- 2:20
km_bag <- matrix(NA, ncol=3, nrow=length(ks))
for(k in ks){
  km_out <- kmeans(scaled_df[, -which(names(scaled_df) == 'NAME')], centers=k, nstart=20)
  km_bag[k - 1, ] <- c(k, km_out$tot.withinss, km_out$betweenss)
}

png('kmeans_btwss_by_k3.png')
plot(km_bag[, 1], km_bag[, 3])
dev.off()

km_bag
km_out <- kmeans(scaled_df[, -which(names(scaled_df) == 'NAME')], centers=4, nstart=20)
table(km_out$cluster) 
scaled_df[km_out$cluster == 3,] # Los Angeles is its own cluster!

cols4 <- brewer.pal(4, "Set1")
##CHANGE: Keeping LA in but highlighting it to show it is outlying
png('kmeans_4_centers_slope_with_LA_highlighted.png')
par(mfrow=c(2, 2))
#Defining pairs to plot instead of repetition
var_pairs <- list(
  c('married.val_2022', 'married.slope_2022'),
  c('unmarried.val_2022', 'unmarried.slope_2022'),
  c('married.slope_2022', 'unmarried.slope_2022'),
  c('no_curve_married', 'no_curve_unmarried')
)
#Looping through pairs and highlighting LA in green
for (i in 1:length(var_pairs)) {
  pair <- var_pairs[[i]]
  plot_colors <- ifelse(scaled_df$NAME == 'Los Angeles County, California', 'green', cols4[km_out$cluster])
  plot(km_out$centers[, pair[1]], km_out$centers[, pair[2]],
       col=plot_colors, pch=16, main=paste("Plot", i))
  
  #Legend showing the clusters and highlighting LA
  legend("topleft", legend=c(1:4, "LA"), fill=c(cols4, 'green'), pch=16)
}
dev.off()

#Highlighting LA in t-sne plot 
cols <- brewer.pal(4, 'Set1')
for(p in seq(5, 35, by=10)){
  tsne_results <- Rtsne(scaled_df,
                        dims = 2, perplexity = p, verbose = TRUE, max_iter = 2500)
  png(glue('perplex_{p}_tsne_iter2500.png'))
  colors <- ifelse(scaled_df$NAME == "Los Angeles County, California", "green", "grey")
  plot(tsne_results$Y[, 1], tsne_results$Y[, 2], main=glue("Perplexity {p}"),
       col=colors, pch=16)
  legend("topright", legend=c("LA", "Other Counties"), fill=c("green", "grey"))
  dev.off()
}

scaled_df[['cluster']] <- km_out$cluster
write.csv(scaled_df, 'cluster_out.csv')


target_vars <- list(c('married.slope_2022', 'B11002_003E'),
                    c('unmarried.slope_2022', 'B11002_012E'))
for(vs in target_vars){
  target_var <- vs[2]
  rep_col_var <- vs[1]
  grp = sub('\\..+', '', rep_col_var)
  # see what bad looks like
  rand_county <- sample(scaled_df[scaled_df[, rep_col_var] == 0, 'NAME'], 1)
  sdf <- ts[ts$NAME == rand_county, c('year', target_var)]
  png(glue('eg_{rep_col_var}_curve.png'))
  plot(sdf$year, sdf[[target_var]], main=glue("No best line for {grp} household count\n {rand_county}"))
  dev.off()
}

# Vis the clusters
for(clus in 1:4){
  # get 3 counties
  rand_counties <- sample(scaled_df[km_out$cluster == clus, 'NAME'], 5)
  for(i in seq_along(rand_counties)){
      rc = rand_counties[i]
      sdf <- ts[ts$NAME == rc, c('year', 'B11002_003E', 'B11002_012E')]
      y1 <- sdf[['B11002_003E']]
      y1 <- (y1 - min(y1)) / (max(y1) - min(y1))
      y2 <- sdf[['B11002_012E']]
      y2 <- (y2 - min(y2)) / (max(y2) - min(y2))
      png(glue('eg_cluster{clus}_{i}_curve.png'))
      plot(sdf$year, y1, main=glue("Cluster {clus} household count\n {rc}"), col="blue", pch=16, ylim=c(0, 1))
      points(sdf$year, y2, col="red", pch=16)
      legend("topleft", fill=c("blue", "red"), legend=c("married", "unmarried"), pch=16, title='household count')
      dev.off()
  }
}
# large variation in values

```

